%% For tips on how to write a great abstract, have a look at
%%	-	https://www.cdc.gov/stdconference/2018/How-to-Write-an-Abstract_v4.pdf (presentation, start here)
%%	-	https://users.ece.cmu.edu/~koopman/essays/abstract.html
%%	-	https://search.proquest.com/docview/1417403858
%%  - 	https://www.sciencedirect.com/science/article/pii/S037837821830402X

\begin{abstract}
Queues are among the most ubiquitous data structures in the field of Computing
Science.
With the advent of multiprocessor programming, concurrent queues are at the
core of many concurrent and distributed algorithms.
Limited work related to surveying concurrent queueing algorithms and
replicating findings from their seminal works have been identified.
This work is an experimental study, with the aim of comparing performance
concurrent queues with one another under different levels
of contention, whilst replicating the results of other researchers. 
The production of a benchmarking framework for concurrent queues also forms
part of this study's contributions. 
We provide evidence to the fact that non-blocking queues are superior to blocking
ones under high contention, and remain competitive under low contention. We
also expose a number of unreported, yet significant biases in a number of
classical works. A limiting factor of this study is that each algorithm is
benchmarked on consumer-grade hardware, fortunately, the benchmarking framework
is developed to run on multiple machines, making this limitation transient in
scope.
Unbounded Multiple Producer Multiple Consumer (MPMC) queues are the focus of
this study.
\end{abstract}